---
title: "Exploring Annotation"
author: "Sifan Zhang"
date: "2024-12-09"
output: html_document
---

**In this workflow we will compare unsupervised annotation with reference-based annotation**
**For reference annotation we will use Azimuth pipeline**
**guidelines can be found in [Azimuth annotation](https://satijalab.github.io/azimuth/articles/run_azimuth_tutorial.html)**

# Prepare  r packages
"azimuth" and "SeuratData" need to be downloaded from github, it's for reference annonation. Downloading could pop up errors, just keep trying installing packages that's missing. 

`devtools::install_github("satijalab/seurat", "seurat5")`
`devtools::install_github("satijalab/seurat-data", "seurat5")`
`remotes::install_github('satijalab/azimuth', ref = 'master')`

```{r libraries, warning=FALSE, message = FALSE}
library(devtools)
library(remotes)
library(SeuratData)
library(SeuratWrappers)
library(Azimuth)
library(SeuratObject)
library(Seurat)
library(Matrix)
library(ggplot2)
library(cowplot)
library(dplyr)
library(tidyr)
library(RColorBrewer)
library(reshape2)
library(pheatmap)
```


# Set up R environment
including working directory, this depends on your own laptop.
For MacOs users you can just copy and paste working directory, for Windows users you need to change backslash to forward slash
```{r setup}
knitr::opts_chunk$set(echo = TRUE, message=F, warning=F)

# sets the directory of all chunks to the directory of this script
knitr::opts_knit$set(root.dir = "M:/SifanZ/OneDrive - Westmead Institute for Medical Research/Data/scRNA_LN/GSE254435")

theme_set(theme_classic())
set.seed(1337)
```

# Load a integerated seurat object, use dot plot to visualize marker gene expression

```{r dataloading}
data <- readRDS("data/integrated.rds")

DefaultAssay(data) <- "RNA"
Markers <- c('CD3E','CD4','CD8A','SELL','CCR7','TCF7','GATA3','CCR4','CXCR3','KLRG1','RORC','CCR6','CXCR5','PDCD1','ICOS',
            'CD40LG','FOXP3','IL2RA','OAS1','ISG15','IFI44L','GZMA','GZMB','GZMH','GZMM','TRDV2','TRAV1-2','KLRB1','SLC4A10',
            'NKG7', 'FCGR3A','NCAM1','CD19','MS4A1','IGHM','FCRL5','IRF4','PRDM1','CD38','CST3','LYZ','CD14','CLEC4C','IL3RA',
            'XCR1', 'CLEC9A','CD1C','FCER1A','IL7R','IL18R1','IL23R','CD34','MKI67')
DotPlot(data, features = Markers, dot.min = 0, dot.scale = 2) +coord_flip() + 
  theme(axis.text = element_text(size = 5)) 
new.cluster.ids <- c(`0` = "Th2_CD4_Tcm", 
                     `1` = "Th17_CD4_Tcm",
                     `2` = "CD14_Mono", 
                     `3` = "CD4_Tcm", 
                     `4` = "CD16_NK",
                     `5` = "IgM_B_mem",
                     `6` = "switched_B_mem",
                     `7` = "Treg_CD4", 
                     `8` = "CD8_cytotoxic", 
                     `9` = "Th2_CD4_Tcm",
                     `10` = "CD8_Tcm",
                     `11` = "vd2_gdT",
                     `12` = "MAIT",
                     `13` = "CD56_NK",
                     `14` = "CD14_Mono",
                     `15` = "CD16_Mono",
                     `16` = "Th2_CD4_Tcm",
                     `17` = "doublets",
                     `18` = "ISG_CD4",
                     `19` = "Tfh_CD4",
                     `20` = "pDC",
                     `21` = "Mixed_T_NK",
                     `22` = "cDC1_cDC2",
                     `23` = "B_Naive",
                     `24` = "cycling",
                     `25` = "ILC",
                     `26` = "CD56low_CD16_NK",
                     `27` = "CXCR3_CD4_Tcm",
                     `28` = "plasmablast",
                     `29` = "HSPC",
                     `30` = "FCRL5_B_mem"
                     )
names(new.cluster.ids) <- levels(data)
data <- RenameIdents(data, new.cluster.ids)
DimPlot(data, reduction = "umap", label = TRUE, label.size = 2.5) + NoLegend()
head(data@meta.data)
```

# Subcluster DCs

Note that at this point DCs can't be separated into cDC1 and cDC2, so we need to do subclustering

```{r subcluster}
sub.cells <- subset(data, idents = "cDC1_cDC2")
DefaultAssay(sub.cells)<-"RNA"
f = "obj.Rdata"
if(!file.exists(f)){
  sub.cells = sub.cells %>%
  NormalizeData() %>%
  FindVariableFeatures() %>%
  ScaleData(features = c('CLEC9A','XCR1','FCER1A','CD1C')) %>%
  RunPCA(features = c('CLEC9A','XCR1','FCER1A','CD1C'))  %>%
  FindNeighbors(dims = 1:45) %>%
  FindClusters(resolution = 0.5) %>%
  RunUMAP(dims = 1:45) 
  save(sub.cells,file = f)
}
load(f)
DimPlot(sub.cells, reduction = 'umap',label = T)+NoLegend()
sub.cells

#Visualize marker genes
DefaultAssay(sub.cells)<-"RNA"
DC_markers<-c('CLEC9A','XCR1','FCER1A','CD1C')
DotPlot(sub.cells,features = DC_markers) + RotatedAxis()

#Annotate and put back to big umap
sub.cells@meta.data$annotation <- ifelse(sub.cells@meta.data$seurat_clusters == 0, "cDC2", "cDC1")
head(sub.cells@meta.data)
data@meta.data$annotation = as.character(Idents(data))
data@meta.data$annotation = ifelse(data@meta.data$annotation=="cDC1_cDC2",
       sub.cells@meta.data$annotation[match(colnames(data@meta.data),colnames(sub.cells@meta.data))],
       data@meta.data$annotation)
Idents(data) = data@meta.data$annotation
DimPlot(data,reduction = "umap", label = T, label.size = 2.5)
table(data@meta.data$annotation)

DotPlot(data, features = Markers, dot.min = 0, dot.scale = 2) +coord_flip() + 
  theme(axis.text = element_text(size = 5)) + theme(axis.text.x = element_text(angle = 45, hjust = 1))


saveRDS(data,"data/alldata_annotated.rds")
```

# Compare Azimuth annotation with unsupervised annotation
If the internet connection is poor, you may need to manually download reference to working directory and install it.
`install.packages("tonsilref.SeuratData_2.0.0.tar.gz", repos = NULL, type = "source")`
Since I have two sample type (blood and lymph node), I tried two reference, PBMC and tonsil.

# PBMC reference

```{r Azimuth PBMC}
data<-readRDS("data/alldata_annotated.rds")

DefaultAssay(data)<-"RNA"
data_pbmc_ref<-RunAzimuth(data, reference = "pbmcref")
head(data_pbmc_ref@meta.data)
data_pbmc_ref
p1<-DimPlot(data_pbmc_ref, reduction = "umap", group.by = "predicted.celltype.l2", label = TRUE, label.size = 2.5) + 
  NoLegend() + theme(aspect.ratio = 1) 
p2<-DimPlot(data, reduction = "umap", label = T, label.size = 2.5) +ggtitle("unsupervised annotation") + 
  NoLegend() + theme(aspect.ratio = 1)
p1+p2

meta.data1 <- data_pbmc_ref@meta.data[, c("annotation", "predicted.celltype.l2")]
meta.data1 <- meta.data1 %>%
  rename(pbmc.predicted.celltype = predicted.celltype.l2)

metadata_long1 <- meta.data1 %>%
  pivot_longer(cols = c("annotation", "pbmc.predicted.celltype"), 
               names_to = "Type", values_to = "CellType") 

counts1 <- metadata_long1 %>%
  group_by(Type, CellType) %>%
  summarise(Count = n(), .groups = "drop")

ggplot(counts1, aes(x = Type, y = Count, fill = CellType)) +
  geom_bar(stat = "identity", position = "stack") +
  theme_minimal() +
  labs(x = "Annotation Type", y = "Number of Barcodes", fill = "Cell Type", title = "Mapping of Cell Types") +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    legend.text = element_text(size = 5),  # Adjust legend text size
    legend.title = element_text(size = 10),  # Adjust legend title size
    legend.key.size = unit(0.5, "cm"),      # Adjust legend key size
    legend.position = "right",              # Position legend to the right
    plot.margin = margin(10, 10, 10, 10),   # Adjust plot margins for more space
    axis.title = element_text(size = 12),   # Increase axis title size (optional)
    axis.text = element_text(size = 10)     # Adjust axis text size (optional)
  )

table(data_pbmc_ref@meta.data$annotation)
table(data_pbmc_ref@meta.data$predicted.celltype.l2)

heatmap_data <- dcast(data_pbmc_ref@meta.data, annotation ~ predicted.celltype.l2, value.var = "predicted.celltype.l2.score", fun.aggregate = mean, na.rm = TRUE)
rownames(heatmap_data) <- heatmap_data$annotation
cell_scores_matrix <- as.matrix(heatmap_data[, -1])
cell_scores_matrix[is.na(cell_scores_matrix)] <- 0
head(cell_scores_matrix)
pheatmap(cell_scores_matrix, 
         cluster_rows = TRUE,     # Cluster cell types (rows)
         cluster_cols = TRUE,     # Cluster conditions (columns)
         scale = "none",          # Optionally scale the data (e.g., "row" for row normalization)
         color = colorRampPalette(c("blue", "white", "red"))(100),  # Adjust color palette
         na_col = "grey",         # Color for missing values (NA)
         main = "Cell Type Score Heatmap",
         fontsize_row = 6,         
         fontsize_col = 10,        
         angle_col = 45 
)

```


# Tonsil

```{r Azimuth Tonsil}
DefaultAssay(data)<-"RNA"
data_tonsil_ref<-RunAzimuth(data, reference = "tonsilref")
head(data_tonsil_ref@meta.data)
data_tonsil_ref
p3<-DimPlot(data_tonsil_ref, reduction = "umap", group.by = "predicted.celltype.l1", label = TRUE, label.size = 2.5) + 
  NoLegend() + theme(aspect.ratio = 1) 
p4<-DimPlot(data, reduction = "umap", label = T, label.size = 2.5) + ggtitle("unsupervised annotation") + 
  NoLegend() + theme(aspect.ratio = 1) 
p3+p4

meta.data2 <- data_tonsil_ref@meta.data[, c("annotation", "predicted.celltype.l1")]
meta.data2 <- meta.data2 %>%
  rename(tonsil.predicted.celltype = predicted.celltype.l1)

metadata_long2 <- meta.data2 %>%
  pivot_longer(cols = c("annotation", "tonsil.predicted.celltype"), 
               names_to = "Type", values_to = "CellType") 

counts2 <- metadata_long2 %>%
  group_by(Type, CellType) %>%
  summarise(Count = n(), .groups = "drop")

ggplot(counts2, aes(x = Type, y = Count, fill = CellType)) +
  geom_bar(stat = "identity", position = "stack") +
  theme_minimal() +
  labs(x = "Annotation Type", y = "Number of Barcodes", fill = "Cell Type", title = "Mapping of Cell Types") +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    legend.text = element_text(size = 5),  # Adjust legend text size
    legend.title = element_text(size = 10),  # Adjust legend title size
    legend.key.size = unit(0.5, "cm"),      # Adjust legend key size
    legend.position = "right",              # Position legend to the right
    plot.margin = margin(10, 10, 10, 10),   # Adjust plot margins for more space
    axis.title = element_text(size = 12),   # Increase axis title size (optional)
    axis.text = element_text(size = 10)     # Adjust axis text size (optional)
  )

table(data_tonsil_ref@meta.data$annotation)
table(data_tonsil_ref@meta.data$predicted.celltype.l1)

heatmap_data <- dcast(data_tonsil_ref@meta.data, annotation ~ predicted.celltype.l1, value.var = "predicted.celltype.l1.score", fun.aggregate = mean, na.rm = TRUE)
rownames(heatmap_data) <- heatmap_data$annotation
cell_scores_matrix <- as.matrix(heatmap_data[, -1])
cell_scores_matrix[is.na(cell_scores_matrix)] <- 0
head(cell_scores_matrix)
pheatmap(cell_scores_matrix, 
         cluster_rows = TRUE,     # Cluster cell types (rows)
         cluster_cols = TRUE,     # Cluster conditions (columns)
         scale = "none",          # Optionally scale the data (e.g., "row" for row normalization)
         color = colorRampPalette(c("blue", "white", "red"))(100),  # Adjust color palette
         na_col = "grey",         # Color for missing values (NA)
         main = "Cell Type Score Heatmap",
         fontsize_row = 6,         
         fontsize_col = 8,        
         angle_col = 45 
)

saveRDS(data_pbmc_ref, "data/data_pbmc_ref.rds")
saveRDS(data_tonsil_ref, "data/data_tonsil_ref.rds")

```

# Annotate DC subclusters using new annotation
I think using PBMC reference gives a better annotation of DC subsets, but I don't want to lose other annotations.
So I will replace DC cells with new annotations, but keep the other cell types unchanged
```{r}
Idents(data_pbmc_ref)<-data_pbmc_ref@meta.data$predicted.celltype.l2
DotPlot(data_pbmc_ref, features = Markers, dot.min = 0, dot.scale = 2) +coord_flip() + 
  theme(axis.text = element_text(size = 5)) + theme(axis.text.x = element_text(angle = 45, hjust = 1)) + ggtitle("predicted cell type")

data_pbmc_ref@meta.data$annotation[
  data_pbmc_ref@meta.data$predicted.celltype.l2 %in% c("cDC1", "cDC2")
] <- data_pbmc_ref@meta.data$predicted.celltype.l2[
  data_pbmc_ref@meta.data$predicted.celltype.l2 %in% c("cDC1", "cDC2")
]

Idents(data_pbmc_ref) <- data_pbmc_ref@meta.data$annotation
Idents(data_pbmc_ref) <- factor(Idents(data_pbmc_ref), levels = c("CD4_Tcm", 
                                                                  "Th2_CD4_Tcm", 
                                                                  "CXCR3_CD4_Tcm",
                                                                  "Th17_CD4_Tcm",
                                                                  "Tfh_CD4",
                                                                  "Treg_CD4",
                                                                  "ISG_CD4",
                                                                  "CD8_Tcm",
                                                                  "CD8_cytotoxic",
                                                                  "vd2_gdT",
                                                                  "MAIT",
                                                                  "CD16_NK",
                                                                  "CD56_NK",
                                                                  "CD56low_CD16_NK",
                                                                  "Mixed_T_NK",
                                                                  "IgM_B_mem",
                                                                  "switched_B_mem",
                                                                  "B_Naive",
                                                                  "FCRL5_B_mem",
                                                                  "plasmablast",
                                                                  "CD14_Mono",
                                                                  "CD16_Mono",
                                                                  "pDC",
                                                                  "cDC1",
                                                                  "cDC2",
                                                                  "ILC",
                                                                  "HSPC",
                                                                  "cycling",
                                                                  "doublets"))


DotPlot(data_pbmc_ref, features = Markers, dot.min = 0, dot.scale = 2.5) +coord_flip() + 
  theme(axis.text = element_text(size = 5)) + theme(axis.text.x = element_text(angle = 45, hjust = 1)) + ggtitle("unsupervised annotation")

p5<-DimPlot(data_pbmc_ref, reduction = "umap", group.by = "predicted.celltype.l2", label = TRUE, label.size = 2.5) + 
  NoLegend() + theme(aspect.ratio = 1) 
p6<-DimPlot(data_pbmc_ref, reduction = "umap", group.by = "annotation", label = TRUE, label.size = 2.5) + 
  NoLegend() + theme(aspect.ratio = 1) + ggtitle("unsupervised annotation")
p5+p6

```

# Session info
```{r session info}
sessionInfo()
```